---
title: "Evidence for strain-specific immunity in bighorn sheep"
author: "Frances Cassirer, Kezia Manlove, Raina Plowright, Tom Besser"
date: "10/10/2016"
output: html_document
---

This file contains the code to recreate Figure 1 and corresponding analyses presented in Cassirer et al. 2016, Evidence for strain-specific immunity in bighorn sheep, Journal of Wildlife Management. 

### Bring in necessary packages
```{r, message = F}
require(survival)
require(lattice)
```

### Read in data
```{r}
bb <- read.csv("~/work/Kezia/Research/FieldStudies/HCBighornContactStudies/Data/RelocationData/BlackButteLocationsFinal_20July2014.csv", header = T, sep = "\t")

bb$UTMN <- bb$Longitude
bb$UTME <- bb$Latitude
bb$DATE <- bb$Date

#-- pull out Uncollared nonunicorn --#
bb <- subset(bb, !(EWEID == ""))
bb$EWEID <- factor(bb$EWEID)
```

## Figure 1 
### Figure 1, Row 1: 2013 observation and symptom stripplots  

```{r, fig.height = 2, fig.width = 6}
#-- BB 2013 --#
health.color2 <- c("black", rgb(1, 1, 1, alpha = 0), "black", 
                   rgb(1, 1, 1, alpha = 0), rgb(1, 1, 1, alpha = 0), "black")

lamb.health.color2 <- c(rgb(1, 1, 1, alpha = 0), "black", "black")

bb2013 <- subset(bb, Year == 2013)
bb2013$EWEID <- factor(bb2013$EWEID)

# Cut down to just Heller Bar ewes

heller.ewes <- c("06BB10", "06BB12", "06BB13", "11LK36")
bb2013.heller <- subset(bb2013, as.character(EWEID) %in% heller.ewes)
bb2013.heller$EWEID <- factor(bb2013.heller$EWEID)

bb3 <- read.csv("./Data/HellerBar_alldata_May_Oct_2013_2014.csv", header = T)
bb3 <- subset(bb3, Year == 2013 & AnimalID %in% heller.ewes)
bb3$EWEID <- factor(bb3$AnimalID, levels = heller.ewes)

bb2013.lambobs <- subset(bb2013.heller, Lamb.Status %in% 
                           c("Lamb Bedded", 
                             "Lamb Bedded Together", 
                             "Lamb Body Contact", 
                             "Lamb Nursed"))

#svg("./Plots/FullSymptomsPlot2013_20160216.svg", height = 1.8, width = 8)
par(mfrow = c(1, 2), oma = c(1, 1, 0, 0), mar = c(4, 6, 1, 1))
plot(as.numeric(factor(bb2013.heller$EWEID)) ~ 
       as.numeric(as.character(bb2013.heller$JulianDate)), 
     cex = 1.2, pch = 16, ylim = c(1, 4), 
     col = health.color2[bb2013.heller$Ewe.Health], 
     xlab = "Julian Date", 
     yaxt = "n", ylab = "", xlim = c(121, 270), xaxt = "n")
points(as.numeric(bb3$EWEID) ~ as.numeric(as.character(bb3$JulianDate)), 
     cex = .5, pch = 1, col = "grey60")
points(as.numeric(factor(bb2013.heller$EWEID)) ~ 
         as.numeric(as.character(bb2013.heller$JulianDate)), 
       cex = 1.2, pch = 16,
       col = health.color2[bb2013.heller$Ewe.Health])
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
axis(side = 2, line = 1, outer = F, las = 1, at = seq(1:4), 
     labels = levels(factor(bb2013.heller$EWEID)))

plot(as.numeric(factor(bb2013.lambobs$EWEID)) ~ 
       as.numeric(as.character(bb2013.lambobs$JulianDate)), 
     cex = 1, pch = 16, col = "white", xlab = "Julian Date", 
     yaxt = "n", ylab = "", xlim = c(121, 270), xaxt = "n")
points(as.numeric(factor(bb2013.lambobs$EWEID)) ~ 
         as.numeric(as.character(bb2013.lambobs$JulianDate)), 
       cex = ifelse(bb2013.lambobs$Symptomatic.Lamb == 2, 
                    0, 
                    .5), 
       pch = 1, 
       col = "grey60")
points(as.numeric(factor(bb2013.lambobs$EWEID)) ~ 
         as.numeric(as.character(bb2013.lambobs$JulianDate)), 
       cex = ifelse(bb2013.lambobs$Symptomatic.Lamb == 1, 
                    1.2, 
                    0), 
       pch = 16, 
       col = ifelse(bb2013.lambobs$Symptomatic.Lamb == 1, 
                    "black", 
                    "white"))
axis(side = 2, line = 1, outer = F, las = 1, at = seq(1:4), 
     labels = levels(factor(bb2013.heller$EWEID)))
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))

polygon(x = c(137, 137, 191, 191), y = c(8.75, 9.25, 9.25, 8.75), col = "grey80", 
        border = "grey80")
#text(x = 141, y = 9, "Ewe Died")
polygon(x = c(120, 120, 140, 140), y = c(0.75, 1.25, 1.25, 0.75), col = "grey80", 
         border = "grey80")
polygon(x = c(181, 181, 270, 270), y = c(0.75, 1.25, 1.25, 0.75), col = "grey80", 
        border = "grey80")

polygon(x = c(120, 120, 133, 133), y = c(1.75, 2.25, 2.25, 1.75), col = "grey80", 
        border = "grey80")
polygon(x = c(179, 179, 270, 270), y = c(1.75, 2.25, 2.25, 1.75), col = "grey80", 
        border = "grey80")

polygon(x = c(120, 120, 137, 137), y = c(2.75, 3.25, 3.25, 2.75), col = "grey80", 
        border = "grey80")
polygon(x = c(163, 163, 270, 270), y = c(2.75, 3.25, 3.25, 2.75), col = "grey80", 
        border = "grey80")

polygon(x = c(120, 120, 125, 125), y = c(3.75, 4.25, 4.25, 3.75), col = "grey80", 
        border = "grey80")
polygon(x = c(180, 180, 270, 270), y = c(3.75, 4.25, 4.25, 3.75), col = "grey80", 
        border = "grey80")

mtext(side = 2, line = 0, outer = T, "Ewe ID")
#dev.off()

```

### Build 2014 observation and symptom stripplot
```{r}
#-- 2014 stripplot --#
bb2014 <- subset(bb, Year == 2014)
bb2014$EWEID <- factor(bb2014$EWEID)
bb2 <- read.csv("./Data/HellerBar_alldata_May_Oct_2013_2014.csv", header = T)
bb2 <- subset(bb2, JulianDate >= 190 & Year == 2014 & AnimalID != "" & 
                AnimalID != ".")
bb2$AnimalID <- ifelse(as.character(bb2$AnimalID) == "14BB53", "14AS52", 
                       as.character(bb2$AnimalID))
bb2$EWEID <- factor(bb2$AnimalID, levels = levels(factor(bb2014$EWEID)))

new.symp.dates <- c(192, 192, 192, 194, 195, 202, 211, 212)
new.symp.animal.ids <- c("06BB13", "11LK36", "14BB03", "06BB12", "06BB10",
                    "14BB04", "14BB02", "14BB02")
new.symp.animal.ids <- c(3, 4, 7, 2, 1, 8, 6, 6)

# svg("./Plots/FullSymptomsPlot_20160216.svg", height = 3, width = 8)
par(mfrow = c(1, 2), oma = c(1, 1, 0, 0), mar = c(4, 6, 1, 1))
plot(as.numeric(factor(bb2014$EWEID)) ~ as.numeric(as.character(bb2014$JulianDate)), 
     cex = .5, pch = 1, col = "grey60", 
     xlab = "Julian Date", yaxt = "n", ylab = "", xlim = c(121, 270), xaxt = "n")
points(as.numeric(bb2$EWEID) ~ as.numeric(as.character(bb2$JulianDate)),  
       cex = .5, pch = 1, col = "grey60")

points(as.numeric(factor(bb2014$EWEID)) ~ as.numeric(as.character(bb2014$JulianDate)), 
     cex = 1.2, pch = 16, col = health.color2[bb2014$Ewe.Health])

# Add Frances's extra points manually
points(new.symp.animal.ids ~ new.symp.dates, pch = 16, col = "black", cex = 1.2)
axis(side = 2, line = 1, outer = F, las = 1, at = seq(1:12), 
     labels = levels(factor(bb2014$EWEID)))
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
polygon(x = c(137, 137, 270, 270), y = c(8.75, 9.25, 9.25, 8.75), 
        col = "grey80", border = "grey80")
polygon(x = c(178, 178, 270, 270), y = c(10.75, 11.25, 11.25, 10.75), 
        col = "grey80", border = "grey80")
polygon(x = c(197, 197, 270, 270), y = c(0.75, 1.25, 1.25, 0.75), 
        col = "grey80", border = "grey80")
polygon(x = c(208, 208, 270, 270), y = c(7.75, 8.25, 8.25, 7.75), 
        col = "grey80", border = "grey80")
polygon(x = c(215, 215, 270, 270), y = c(5.75, 6.25, 6.25, 5.75), 
        col = "grey80", border = "grey80")
mtext(side = 2, line = 0, outer = T, "Ewe ID")

bb2014.lambobs <- subset(bb2014, Lamb.Status %in% 
                           c("Lamb Bedded", 
                             "Lamb Bedded Together", 
                             "Lamb Body Contact", 
                             "Lamb Nursed"))
plot(as.numeric(bb2014.lambobs$EWEID) ~ 
       as.numeric(as.character(bb2014.lambobs$JulianDate)),
     cex = ifelse(bb2014.lambobs$Symptomatic.Lamb == 2, 
                  0, 
                  bb2014.lambobs$Symptomatic.Lamb * 1.2), 
     pch = 16, col = health.color2[bb2014.lambobs$Ewe.Health], 
     xlab = "Julian Date", 
     yaxt = "n", ylab = "", 
     xlim = c(121, 270), xaxt = "n", ylim = c(1, 12))
points(as.numeric(bb2014.lambobs$EWEID) ~ 
         as.numeric(as.character(bb2014.lambobs$JulianDate)),
       cex = .5, 
       pch = 1, col = "grey60")
points(as.numeric(bb2014.lambobs$EWEID) ~ as.numeric(as.character(bb2014.lambobs$JulianDate)),
     cex = ifelse(bb2014.lambobs$Symptomatic.Lamb == 2, 
                  0, 
                  bb2014.lambobs$Symptomatic.Lamb * 1.2), 
     pch = 16, col = health.color2[bb2014.lambobs$Ewe.Health])
points(as.numeric(bb2014.lambobs$EWEID) ~ 
         as.numeric(as.character(bb2014.lambobs$JulianDate)), 
       cex = ifelse(bb2014.lambobs$Symptomatic.Lamb == 2, 0, 
                    bb2014.lambobs$Symptomatic.Lamb * 1.2), pch = 16, 
       col = lamb.health.color2[as.numeric(bb2014.lambobs$Symptomatic.Lamb) + 1])

points(as.numeric(bb2014.lambobs$EWEID) ~ 
         as.numeric(as.character(bb2014.lambobs$JulianDate)),
     cex = ifelse(bb2014.lambobs$Symptomatic.Lamb == 2, 
                  0, 
                  bb2014.lambobs$Symptomatic.Lamb * 1.2), 
     pch = 16, col = health.color2[bb2014.lambobs$Ewe.Health])
axis(side = 2, line = 1, outer = F, las = 1, at = seq(1:12), 
     labels = levels(factor(bb2014$EWEID)))
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
polygon(x = c(137, 137, 270, 270), y = c(8.75, 9.25, 9.25, 8.75), 
        col = "grey80", border = "grey80")
polygon(x = c(178, 178, 270, 270), y = c(10.75, 11.25, 11.25, 10.75), 
        col = "grey80", border = "grey80")
mtext(side = 2, line = 0, outer = T, "Ewe ID")
# 06BB10
polygon(x = c(120, 120, 137, 137), y = c(.75, 1.25, 1.25, .75), 
        col = "grey80", border = "grey80")
polygon(x = c(139, 139, 270, 270), y = c(.75, 1.25, 1.25, .75), 
        col = "grey80", border = "grey80")
# 06BB12
polygon(x = c(183, 183, 270, 270), y = c(1.75, 2.25, 2.25, 1.75), 
        col = "grey80", border = "grey80")
# 06BB13
polygon(x = c(120, 120, 132, 132), y = c(2.75, 3.25, 3.25, 2.75), 
        col = "grey80", border = "grey80")
polygon(x = c(177, 177, 270, 270), y = c(2.75, 3.25, 3.25, 2.75), 
        col = "grey80", border = "grey80")
# 11LK36
polygon(x = c(120, 120, 132, 132), y = c(3.75, 4.25, 4.25, 3.75), 
        col = "grey80", border = "grey80")
polygon(x = c(178, 178, 270, 270), y = c(3.75, 4.25, 4.25, 3.75), 
        col = "grey80", border = "grey80")
# 14AS52
polygon(x = c(120, 120, 125, 125), y = c(4.75, 5.25, 5.25, 4.75), 
        col = "grey80", border = "grey80")
polygon(x = c(163, 163, 270, 270), y = c(4.75, 5.25, 5.25, 4.75), 
        col = "grey80", border = "grey80")
# 14BB02
polygon(x = c(120, 120, 133, 133), y = c(5.75, 6.25, 6.25, 5.75), 
        col = "grey80", border = "grey80")
polygon(x = c(167, 167, 270, 270), y = c(5.75, 6.25, 6.25, 5.75), 
        col = "grey80", border = "grey80")
# 14BB03
polygon(x = c(177, 177, 270, 270), y = c(6.75, 7.25, 7.25, 6.75), 
        col = "grey80", border = "grey80")
# 14BB04
polygon(x = c(120, 120, 131, 131), y = c(7.75, 8.25, 8.25, 7.75), col = "grey80", border = "grey80")
polygon(x = c(183, 183, 270, 270), y = c(7.75, 8.25, 8.25, 7.75), col = "grey80", border = "grey80")
# 14BB05
polygon(x = c(120, 120, 270, 270), y = c(8.75, 9.25, 9.25, 8.75), col = "grey80", border = "grey80")
# 14BB07
polygon(x = c(120, 120, 125, 125), y = c(9.75, 10.25, 10.25, 9.75), col = "grey80", border = "grey80")
polygon(x = c(173, 173, 270, 270), y = c(9.75, 10.25, 10.25, 9.75), col = "grey80", border = "grey80")
# 14BB09
polygon(x = c(120, 120, 123, 123), y = c(10.75, 11.25, 11.25, 10.75), col = "grey80", border = "grey80")
polygon(x = c(145, 145, 270, 270), y = c(10.75, 11.25, 11.25, 10.75), col = "grey80", border = "grey80")
# 14BB17
polygon(x = c(120, 120, 125, 125), y = c(11.75, 12.25, 12.25, 11.75), col = "grey80", border = "grey80")
polygon(x = c(163, 163, 270, 270), y = c(11.75, 12.25, 12.25, 11.75), col = "grey80", border = "grey80")
#dev.off()
```

### Figure 1, Row 2: Survival curves
Read in data. 
```{r}
bb <- read.csv("./Data/HellerBar_alldata_May_Oct_2013_2014.csv", header = T)

bb$UTMN <- bb$Longitude
bb$UTME <- bb$Latitude
bb$DATE <- bb$Date
bb$EWEID <- bb$AnimalID

# pull out Uncollared nonunicorn
bb <- subset(bb, !(EWEID == "")) 
   # subsets bb dataset so that anybody with EWEID == "" gets removed
bb$EWEID <- factor(bb$EWEID) 
   # rewrites the EWEID variable as a categorical variable

# separate 2013 and 2014 data
bb2014 <- subset(bb, Year == 2014)
bb2014$EWEID <- factor(bb2014$EWEID)

bb2013 <- subset(bb, Year == 2013)
bb2013$EWEID <- factor(bb2013$EWEID)
```
Mortality hazards for ewes and lambs over summer

```{r}
# tabulate number at risk and number events at each death date
ad.time1 <- rep(120, 13)
  # study starts with 13 ewes on May 1 (Julian date 120)
#ad.time2 <- c(137, 178, 197, 215, 208, 215, rep(275, 7))
ad.time2 <- c(137, 178, 197, 208, 215, rep(275, 8))
# these mortality dates are pulled from notes on ewe mortality dates
  # not included in bb location data per se. 275 is Julian date for October 1
ad.event <- c(rep(1, 5), rep(0, 8))
  # first 5 end dates in ad.time2 are deaths ("events"), last 8 are censored
ad.data <- as.data.frame(cbind(ad.time1, ad.time2, ad.event))
  # bind the start (ad.time1), end (ad.time2), and event status (ad.event)
  # vectors together into a dataframe named ad.data

# tabulate number lambs at risk, and number events at each lamb death date
lamb.time1 <- c(117, 117, 123, 125, 125, 125, 131, 132, 132, 132, 137)
lamb.time2 <- c(183, 178, 145, 173, 163, 163, 183, 177, 175, 167, 139)

lamb.event <- rep(1, length(lamb.time1))
lamb.data <- as.data.frame(cbind(lamb.time1, lamb.time2, lamb.event))

# follow the same basic protocol to build survival data 
# (start time, stop time, event status) for 2004-2013.
  # adults  @ Heller Bar 2004-2013
  # 06BB06 # first summer 2006; dls 8/22/2007; 1 full summer
  # 06BB10 # first summer 2006; 8 full summers
  # 06BB11 # first summer 2006; dls 07/01/08; 2 full summers
  # 06BB12 # first summer 2006; 8 full summers
  # 06BB13 # first summer 2006; 8 full summers

# ALL 2004-2013 ewes
ad.time1.2013 <- rep(120, 29)
ad.time2.2013 <- c(182, 230, rep(270, 27))
ad.event.2013 <- c(rep(1, 2), rep(0, 27))
ad.data.2013 <- as.data.frame(cbind(ad.time1.2013, ad.time2.2013, ad.event.2013))

# ONLY 2013 ewes
ad.time1.2013 <- rep(120, 4)
ad.time2.2013 <- rep(270, 4)
ad.event.2013 <- rep(0, 4)
ad.data.2013 <- as.data.frame(cbind(ad.time1.2013, ad.time2.2013, ad.event.2013))

# build lamb survival data @ Heller Bar 2004-2013
# pull in my own 2013 lamb survival data
lamb.time1.2013 <- c(140, 140, 137, 125)
lamb.time2.2013 <- c(181, 179, 163, 180)
lamb.event.2013 <- rep(1, length(lamb.time1.2013))
lamb.data.2013 <- as.data.frame(cbind(lamb.time1.2013, lamb.time2.2013, lamb.event.2013))
# add in data from 2004-2012 from merged lamb data
# read in merged lamb data
merged.lambs <- read.csv("./Data/MergedLambData_100311_netupdate_revisedForBB11And12.csv", header = T)
merged.heller <- subset(merged.lambs, EWEID %in% c("06BB06", "06BB10", "06BB12", 
                                                   "06BB11", "06BB12", "06BB13"))
merged.heller$Julian.birth <- c(125, 129, 129, 136, 136, 129, 129, 129, 117, 
                                128, 136, 129, 131, 138, 121, 127, 133, 124, 
                                125, 141, NA, 125, 134)
pre.2013.lamb.morts <- merged.heller$Julian.birth + merged.heller$SURV_DAYS
lamb.time1.2013 <- c(lamb.time1.2013, merged.heller$Julian.birth[c(-21)])
lamb.time2.2013 <- c(lamb.time2.2013, pre.2013.lamb.morts[c(-21)])
lamb.event.2013 <- c(lamb.event.2013, merged.heller$CENSOR2[c(-21)])
lamb.data.2013 <- as.data.frame(cbind(lamb.time1.2013, lamb.time2.2013, lamb.event.2013))

# # build histograms of mort date for 2014 and 2004-2013
# par(mfrow = c(2, 1))
# breaks.in <- seq(120, 280, length.out = 17)
# hist(lamb.time2.2013, xlim = c(120, 240), breaks = breaks.in, 
#      xlab = "Julian date of lamb mortalities", main = "2004-2013", col = "grey60")
# hist(lamb.time2, xlim = c(120, 240), breaks = breaks.in, 
#      xlab = "Julian date of lamb mortalities", main = "2014", col = "grey60")
```

### Figure 1, Row 3: Ewe and lamb mortality hazards
Cox PH model for adults/lambs

```{r}
# 1. pull adult survival data from 2013 and 2014 into one dataframe to plug into
  # Cox proportional hazards model
ad.full.time1 <- c(ad.time1.2013, ad.time1)
ad.full.time2 <- c(ad.time2.2013, ad.time2)
ad.full.event <- c(ad.event.2013, ad.event)
  # build an indicator variable for year (2004-2013 = 0; 2014 = 1)
ad.full.year <- c(rep(0, length(ad.event.2013)), rep(1, length(ad.event)))

  # 2. use survfit function to tabulate adult data to build Kaplan-Meier curves below
ad.survfit <- survfit(Surv(ad.full.time1, ad.full.time2, ad.full.event) ~ factor(ad.full.year))
ad.survfit.2013 <- survfit(Surv(ad.time1.2013, ad.time2.2013, ad.event.2013) ~ 1)
ad.survfit.2014 <- survfit(Surv(ad.time1, ad.time2, ad.event) ~ 1)

  # 3. fit Cox proportional hazards model for adults with "ad.full.year" as a predictor
coxph.ad <- coxph(Surv(ad.full.time1, ad.full.time2, ad.full.event) ~ factor(ad.full.year))
summary(coxph.ad) 
  # results from log-rank test are in the row with column labels "coef", "exp(coef)",
  # "se(coef)", "z", "Pr(>|z|)". The last one is the p-value. 

# combine 2014 and 2004-2013 lamb data and build Cox PH model for lambs
lamb.full.time1 <- c(lamb.time1.2013, lamb.time1)
lamb.full.time2 <- c(lamb.time2.2013, lamb.time2)
lamb.full.event <- c(lamb.event.2013, lamb.event)
lamb.full.year <- c(rep(0, length(lamb.event.2013)), rep(1, length(lamb.event)))

lamb.survfit <- survfit(Surv(lamb.full.time1, lamb.full.time2, lamb.full.event) ~ factor(lamb.full.year))
lamb.survfit.2013 <- survfit(Surv(lamb.time1.2013, lamb.time2.2013, lamb.event.2013) ~ 1)
lamb.survfit.2014 <- survfit(Surv(lamb.time1, lamb.time2, lamb.event) ~ 1)

coxph.lambs <- coxph(Surv(lamb.full.time1, lamb.full.time2, lamb.full.event) ~ factor(lamb.full.year))
summary(coxph.lambs)
```

### Plot Kaplan-Meier curves for adults and lambs, 2014 and 2004-2013

```{r, fig.height = 4, fig.width = 6}
# svg("./Plots/KaplanMeierCurves_20160130.svg", width = 8, height = 4)
par(mfrow = c(1, 2)) # specify plot is two panels in one row
# build adult Kaplan-Meier curves
plot(ad.survfit, conf.int = T, xlab = "Date", ylab = "Survival", 
      xlim = c(120, 250), col = c("grey60", "black"), lwd = c(1, 1), 
     xaxt = "n")
lines(ad.survfit, conf.int = F, xlab = "Date", ylab = "", 
     main = "Ewes", xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))

# build legend and label x-axis with actual dates (as opposed to Julian)
leg.text <- c("2014", "2004-2013")
legend("bottomleft", leg.text, col = c("black", "grey60"), bty = "n", lwd = c(2, 2))
axis(side = 1, at = c(120, 151, 181, 212, 243), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))

# build lamb Kaplan-Meier curves
#svg("./Plots/LambSurvCurvesRevised_20160616.svg", width = 4, height = 5)
plot(lamb.survfit, conf.int = T, xlab = "Date", ylab = "Survival", 
     xlim = c(120, 250), col = c("grey60", "black"), xaxt = "n")
lines(lamb.survfit, conf.int = F, xlab = "Date", ylab = "", 
     xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))
# label x-axis
axis(side = 1, at = c(120, 151, 181, 212, 243), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
# dev.off()
```
Alternative version
```{r, fig.height = 4, fig.width = 6}
par(mfrow = c(1, 2))
plot(lamb.survfit.2013, conf.int = T, xlab = "Date", ylab = "Survival", 
     xlim = c(120, 250), col = c("grey60", "black"), xaxt = "n", main = "2013")
lines(lamb.survfit.2013, conf.int = F, xlab = "Date", ylab = "", 
      xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))
plot(lamb.survfit.2014, conf.int = T, xlab = "Date", ylab = "Survival", 
     xlim = c(120, 250), col = c("grey60", "black"), xaxt = "n", main = "2014")
lines(lamb.survfit.2014, conf.int = F, xlab = "Date", ylab = "", 
      xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))

```

### Figure 1, Row 3. Moving average clinical signs

```{r, fig.height = 4, fig.width = 6}
bb2013$EweClinicalSignsDailyScore <- bb2013$Ewe_Symptomfc

# extract ewe-specific data. all data from each ewe becomes an element in a list
# with as many elements as there are collared ewes (4 in 2013). 
ewe.list.2013 <- vector("list", length(levels(bb2013$EWEID)))
for(i in 1:length(ewe.list.2013)){ # loop over 4 EWEIDs present in 2013
  ewe.list.2013[[i]] <- subset(bb2013, EWEID == levels(bb2013$EWEID)[i])
  # extract all 2013 data associated with each ewe
}

# build matrix that is days x ewes; calculate seven-day moving average for each 
# ewe over study timespan
mov.avg.mat.2013 <- matrix(NA, ncol = range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1], 
                           nrow = length(levels(bb2013$EWEID)))
for(i in 1:dim(mov.avg.mat.2013)[1]){ # loop over each of the four 2013 ewes
  ewe.dat.2013 <- ewe.list.2013[[i]]
  for(j in 1:(range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1] - 6)){
    # loop over each block of 7 days across the timespan of the study
    mov.avg.mat.2013[i, j] <- mean(na.omit(subset(ewe.dat.2013, 
                                                  JulianDate %in% seq(122 + j - 3, 122 + j + 3))$EweClinicalSignsDailyScore))
    # get the mean clinical score for each ewe over each 7-day window
    mov.avg.mat.2013[i, j] <- ifelse(is.na(mov.avg.mat.2013[i, j]) == T, 0, mov.avg.mat.2013[i, j])    
    # replace "NA" clinical scores with "0"'s
  }
}

# get clinical sign scores from 2013 aggregated over all 4 collared ewes
# by summing up each ewe's score for each 7-day window and dividing by 4
aggregate.clin.signs.2013 <- apply(mov.avg.mat.2013, 2, sum) / 4

# 2014: rinse and repeat procedure outlined above. 

bb2014$EweClinicalSignsDailyScore <- bb2014$Ewe_Symptomfc

# extract ewe-specific data
ewe.list <- vector("list", length(levels(bb2014$EWEID)))
for(i in 1:length(ewe.list)){
  ewe.list[[i]] <- subset(bb2014, EWEID == levels(bb2014$EWEID)[i])
}

# build matrix that is days x ewes; calculate seven-day moving average for each ewe over range of study
mov.avg.mat <- matrix(NA, ncol = range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1], 
                      nrow = length(levels(bb2014$EWEID)))
for(i in 1:dim(mov.avg.mat)[1]){
  ewe.dat <- ewe.list[[i]]
  for(j in 1:(range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1] - 6)){
    mov.avg.mat[i, j] <- mean(na.omit(subset(ewe.dat, 
                                             JulianDate %in% seq(122 + j - 3, 122 + j + 3))$EweClinicalSignsDailyScore))
    mov.avg.mat[i, j] <- ifelse(is.na(mov.avg.mat[i, j]) == T, 0, mov.avg.mat[i, j])    
  }
}

aggregate.clin.signs <- apply(mov.avg.mat, 2, sum) / 12

bb2013$LambClinicalSignsDailyScore2 <- bb2013$Lamb_SymptomsFC

# extract ewe-specific data
lamb.list.2013 <- vector("list", length(levels(bb2013$EWEID)))
for(i in 1:length(lamb.list.2013)){
  lamb.list.2013[[i]] <- subset(bb2013, EWEID == levels(bb2013$EWEID)[i])
}

mov.avg.mat.l2.2013 <- matrix(NA, ncol = range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1], 
                              nrow = length(levels(bb2013$EWEID)))
for(i in 1:dim(mov.avg.mat.l2.2013)[1]){
  lamb.dat.2013 <- lamb.list.2013[[i]]
  for(j in 1:(range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1] - 6)){
    mov.avg.mat.l2.2013[i, j] <- mean(na.omit(subset(lamb.dat.2013, 
                                                     JulianDate %in% seq(122 + j - 3, 122 + j + 3))$LambClinicalSignsDailyScore2))
    mov.avg.mat.l2.2013[i, j] <- ifelse(is.na(mov.avg.mat.l2.2013[i, j]) == T, 0, 
                                        mov.avg.mat.l2.2013[i, j])    
  }
}

aggregate.clin.signs.l2.2013 <- apply(mov.avg.mat.l2.2013, 2, sum) / 4

bb2014$LambClinicalSignsDailyScore2 <- bb2014$Lamb_SymptomsFC
  
# extract ewe-specific data
lamb.list <- vector("list", length(levels(bb2014$EWEID)))
for(i in 1:length(lamb.list)){
  lamb.list[[i]] <- subset(bb2014, EWEID == levels(bb2014$EWEID)[i])
}


mov.avg.mat.l2 <- matrix(NA, ncol = range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1], nrow = length(levels(bb2014$EWEID)))
for(i in 1:dim(mov.avg.mat.l2)[1]){
  lamb.dat <- lamb.list[[i]]
  for(j in 1:(range(bb2014$JulianDate)[2] - range(bb2014$JulianDate)[1] - 6)){
    mov.avg.mat.l2[i, j] <- mean(na.omit(subset(lamb.dat, JulianDate %in% seq(122 + j - 3, 122 + j + 3))$LambClinicalSignsDailyScore2))
    mov.avg.mat.l2[i, j] <- ifelse(is.na(mov.avg.mat.l2[i, j]) == T, 0, mov.avg.mat.l2[i, j])    
  }
}

lamb.time1 <- c(117, 117, 123, 125, 125, 125, 131, 132, 132, 132, 137)
lamb.time2 <- c(183, 178, 145, 173, 163, 163, 183, 177, 175, 167, 139)

# lambs per day from 120 to 270
lamb.denom.14 <- c(rep(11, 19), rep(10, 6), rep(9, 18), rep(7, 4), rep(6, 6), rep(5, 2), rep(4, 2), rep(3, 1), rep(2, 5), rep(1, 96))
aggregate.clin.signs.l2 <- apply(mov.avg.mat.l2, 2, sum) / lamb.denom.14

```

### Compiled version of bottom two panels

```{r}
# reordered so that both ewe dynamics are on a single plot, and both lamb dynamics are on the other
# svg("./Plots/BBClinSignSeverity_20160202.svg", height = 6, width = 7)
par(mfrow = c(2, 2), oma = c(2, 2, 1, 2), mar = c(2, 2, 1, 3), xpd = F)
plot(0, 0, col = "black", type = "p", yaxt = "n", ylab = "", xlab = "Date", 
     xaxt = "n", xlim = c(120, 260),  lty = 2, lwd = .75, ylim = c(0, 1.7), 
     pch = 16, cex = .6)
lines(lowess(aggregate.clin.signs.2013[1:153] ~ seq(124, 124 + 152), f = 1/4), 
      col = "black",  lty = 1, lwd = 3)
points(aggregate.clin.signs.2013[1:153] ~ seq(124, 124 + 152), pch = 16, cex = .6)
points(aggregate.clin.signs[1:153] ~ seq(124, 124 + 152), pch = 16, cex = .6, col = "grey70")
lines(lowess(aggregate.clin.signs[1:153] ~ seq(124, 124 + 152), f = 1/4), col = "grey70",  
      lty = 1, lwd = 3)
axis(side = 2, at = seq(0, 1.5, length.out = 4), 
     label = round(seq(0, 1.5, length.out = 4), 2))
mtext(side = 2, outer = F, line = 2, "Severity of clinical signs", las = 0)
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
leg.text <- c("2013", "2014")
legend(x = 180, y = 1.77, leg.text, col = c("black", "grey70"), lwd = c(1, 1), lty = c(1, 1), 
       pch = c(16, 16), pt.cex = c(.6, .6), bty = "n", cex = 1)
mtext(side = 3, "Ewes")

plot(0, 0, ylim = c(0, 1.7), col = "black", type = "p", yaxt = "n", ylab = "", xlab = "", 
     xaxt = "n", xlim = c(120, 260), lty = 2, lwd = .75, pch = 16, cex = .6)
lines(lowess(aggregate.clin.signs.l2.2013[1:153] ~ seq(124, 124 + 152), f = 1/8), 
      col = "black",  lty = 1, lwd = 3)
points(aggregate.clin.signs.l2.2013[1:153] ~ seq(124, 124 + 152), pch = 16, 
       col = "black", lty = 2, cex = .6)
lines(lowess(as.numeric(as.character(aggregate.clin.signs.l2[1:152])) ~ seq(124, 124 + 151), f = 1/20), 
      col = "grey70",  lty = 1, lwd = 3)
points(jitter(aggregate.clin.signs.l2, 0)[1:153] ~ seq(124, 124 + 152), pch = 16, 
       cex = .6, col = "grey70")
axis(side = 2, at = seq(0, 1.5, length.out = 4), 
     label = round(seq(0, 1.5, length.out = 4), 2))
axis(las = 1, side = 1, at = 120 + c(0, 31, 61, 92, 123), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
#mtext(side = 2, outer = F, line = 2, "", las = 0)
leg.text <- c("2013", "2014")
legend(x = 180, y = 1.77, leg.text, col = c("black", "grey70"), lwd = c(1, 1), lty = c(1, 1), 
       pch = c(16, 16), pt.cex = c(.6, .6), bty = "n", cex = 1)
mtext(side = 3, "Lambs")

# Plot Kaplan-Meier curves for adults and lambs, 2014 and 2004-2013
# svg("./Plots/KaplanMeierCurves_20160130.svg", width = 8, height = 4)
# build adult Kaplan-Meier curves
plot(ad.survfit, conf.int = T, xlab = "Date", ylab = "Survival", 
     xlim = c(120, 250), col = c("grey60", "black"), lwd = c(1, 1), 
     xaxt = "n")
lines(ad.survfit, conf.int = F, xlab = "Date", ylab = "", 
      main = "Ewes", xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))

# build legend and label x-axis with actual dates (as opposed to Julian)
leg.text <- c("2014", "2004-2013")
legend("bottomleft", leg.text, col = c("black", "grey60"), bty = "n", lwd = c(2, 2))
axis(side = 1, at = c(120, 151, 181, 212, 243), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
mtext(side = 2, outer = F, line = 2, "Survival", las = 0)

# build lamb Kaplan-Meier curves
plot(lamb.survfit, conf.int = T, xlab = "Date", ylab = "Survival", 
     xlim = c(120, 250), col = c("grey60", "black"), xaxt = "n")
lines(lamb.survfit, conf.int = F, xlab = "Date", ylab = "", 
      xlim = c(120, 250), col = c("grey60", "black"), lwd = c(2, 2))
# label x-axis
axis(side = 1, at = c(120, 151, 181, 212, 243), 
     labels = c("May 1", "June 1", "July 1", "Aug 1", "Sept 1"))
# dev.off()
```

## Figure 2

```{r, echo=FALSE, fig.height = 4, fig.width = 6}
#data <- read.csv("./Data/Movi_Serol_20160202.csv", header = T)

data <- read.csv("./Data/Movi_Serol_20160302.csv", header = T)
before.dat <- subset(data, Time.relative.to.outbreak == "Before")
during.dat <- subset(data, Time.relative.to.outbreak == "During")
after.dat <- subset(data, Time.relative.to.outbreak == "Post")

before.binom.test <- t.test(x = before.dat$Movipneumonia.ELISA, conf.level = .9)$conf.int
during.binom.test <- t.test(x = during.dat$Movipneumonia.ELISA, conf.level = .9)$conf.int
after.binom.test <- t.test(x = after.dat$Movipneumonia.ELISA, conf.level = .9)$conf.int


#svg("./Plots/MoviTiters_V6_20160217.svg", width = 3, height = 4)
par(mfrow = c(1, 2))
plot(data$Movipneumonia.ELISA ~ jitter(as.numeric(data$Time.relative.to.outbreak), .2), 
     type = "p", xaxt = "n", ylab = "ELISA % Inhibition", xlab = "", pch = 1,
     cex = .7, las = 1, col = "grey40",
     ylim = c(40, 100))
axis(side = 1, at = c(1, 2, 3), labels = c("Before", "During", "After"))
segments(x0 = 1, x1 = 1, 
         y0 = before.binom.test[[1]], y1 = before.binom.test[[2]], lwd = 1)
segments(x0 = 2, x1 = 2, 
         y0 = during.binom.test[[1]], y1 = during.binom.test[[2]], lwd = 1)
segments(x0 = 3, x1 = 3, 
         y0 = after.binom.test[[1]], y1 = after.binom.test[[2]], lwd = 1)
segments(x0 = .95, x1 = 1.05, 
         y0 = before.binom.test[[1]], y1 = before.binom.test[[1]], lwd = 1)
segments(x0 = .95, x1 = 1.05, 
         y0 = before.binom.test[[2]], y1 = before.binom.test[[2]], lwd = 1)
segments(x0 = 1.95, x1 = 2.05, 
         y0 = during.binom.test[[1]], y1 = during.binom.test[[1]], lwd = 1)
segments(x0 = 1.95, x1 = 2.05, 
         y0 = during.binom.test[[2]], y1 = during.binom.test[[2]], lwd = 1)
segments(x0 = 2.95, x1 = 3.05, 
         y0 = after.binom.test[[1]], y1 = after.binom.test[[1]], lwd = 1)
segments(x0 = 2.95, x1 = 3.05, 
         y0 = after.binom.test[[2]], y1 = after.binom.test[[2]], lwd = 1)
segments(x0 = 2.975, x1 = 3.025, 
         y0 = mean(after.dat$Movipneumonia.ELISA), y1 = mean(after.dat$Movipneumonia.ELISA), lwd = 1)
segments(x0 = 1.975, x1 = 2.025, 
         y0 = mean(during.dat$Movipneumonia.ELISA), y1 = mean(during.dat$Movipneumonia.ELISA), lwd = 1)
segments(x0 = 0.975, x1 = 1.025, 
         y0 = mean(before.dat$Movipneumonia.ELISA), y1 = mean(before.dat$Movipneumonia.ELISA), lwd = 1)
mtext(side = 1, line = 3, "Time relative to outbreak")
#dev.off()

# Pre-outbreak ELISA for animals that died and survived, respectively. 
survived <- subset(before.dat, Fate == "Survived")
died <- subset(before.dat, Fate == "Died")
survd.before.binom.test <- t.test(x = survived$Movipneumonia.ELISA, conf.level = .9)$conf.int
died.before.binom.test <- t.test(x = died$Movipneumonia.ELISA, conf.level = .9)$conf.int

# svg("./Plots/PreOutbreakELISAByFate_REVISED_20160302.svg",  height = 4, width = 3)
plot(before.dat$Movipneumonia.ELISA ~ as.numeric(before.dat$Fate), 
     type = "p", xaxt = "n", ylab = "ELISA % Inhibition", xlab = "", pch = 1,
     cex = .7, las = 1, col = "grey40", xlim = c(.5, 2.5),
     ylim = c(40, 100))
segments(x0 = 1, x1 = 1, 
         y0 = died.before.binom.test[[1]], y1 = died.before.binom.test[[2]], lwd = 1)
segments(x0 = 2, x1 = 2, 
         y0 = survd.before.binom.test[[1]], y1 = survd.before.binom.test[[2]], lwd = 1)
segments(x0 = .95, x1 = 1.05, 
         y0 = died.before.binom.test[[1]], y1 = died.before.binom.test[[1]], lwd = 1)
segments(x0 = .95, x1 = 1.05, 
         y0 = died.before.binom.test[[2]], y1 = died.before.binom.test[[2]], lwd = 1)
segments(x0 = 1.95, x1 = 2.05, 
         y0 = survd.before.binom.test[[1]], y1 = survd.before.binom.test[[1]], lwd = 1)
segments(x0 = 1.95, x1 = 2.05, 
         y0 = survd.before.binom.test[[2]], y1 = survd.before.binom.test[[2]], lwd = 1)
segments(x0 = 1.975, x1 = 2.025, 
         y0 = mean(survived$Movipneumonia.ELISA), y1 = mean(survived$Movipneumonia.ELISA), lwd = 1)
segments(x0 = 0.975, x1 = 1.025, 
         y0 = mean(died$Movipneumonia.ELISA), y1 = mean(died$Movipneumonia.ELISA), lwd = 1)
text(x = 1, y = 95, "N = 5")
text(x = 2, y = 95, "N = 6")
axis(side = 1, at = c(1, 2), labels = c("Died", "Survived"))
# dev.off()
```

